Verifying Fine-Grained and Lock-Free Binary Search Trees

### 2021-07-08 concurrency/Verification of Concurrent Binary Search Tree.tex
trying to describe generic atomic specs

\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsthm}
%\usepackage{amssymb}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{iris}
%\usepackage{lstlangcoq}
\renewcommand{\lstlistingname}{Figure}

@@

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\renewcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
\newcommand{\us}{$\mu$s}

%\newcommand{\ignore}[1]{}

@@

\title{Verifying a Binary Search Tree with Fine-Grained Locking}
\author{Roshan Sharma}
\author{Shengyi Wang}
\author{Lennart Beringer}
\author{William Mansky}

@@

\catcode`\@=11
\section{Introduction}
\input{intro.tex}
Modern CPU design trends have shifted towards multicore processing over the last two decades. This is because the computing power of computers has reached its limit in single-threaded execution, while the quantity of transistors per single-core CPU are increasing as described by Moore's law \cite{moore}. CPUs have multiple cores now because that is the only way to get more transistors per chip, and achieve a desirable throughput. However, doubling the quantity of cores on one chip does not immediately imply an increase in the performance by a factor of two. Programs that we aim to run in multiple cores have to be carefully designed to efficiently divide the workload into threads, and facilitate non-conflicting access to the shared data by multiple threads in parallel. These challenges is resolved by designing the concurrent data structures, algorithms that provide thread-safe access to the shared data by storing and organizing data in a standard way. Concurrent data structures are now widely employed in real-world applications like database algorithms, file systems, networking, e-commerce, etc. Designing these data structures has been considered as one of the difficult tasks in the software development fields. Despite the careful design, implementation, and thorough unit testing of these algorithms, it is very likely to expect subtle bugs and unexpected behaviors of the program. This is because threads interact with each other in complex and subtle ways such that it is almost impossible to cover every corner case while testing. Therefore, it is important to use a precise way of proving the correctness of concurrent data structures which provides a guarantee that the given algorithm correctly implements the specified concurrent data structures.

Formal verification is an active field of research that aims to prove or disprove the correctness of algorithms with respect to a certain formal specification, using mathematical techniques. Model checking and theorem proving  (also called deductive reasoning) are the two main approaches in this field. Model checking builds the mathematical model of a system as a collection of states and actions that move from one state to another. The model checker then outputs yes if the given model meets the given specifications and produces a counterexample otherwise, after a systematic and exhaustive exploration of a mathematical model. Model checking approach only works for finite models and certain infinite models that can be easily abstracted into finite sets of states. While it is a simple and automated operation, it does not scale well to large and complex systems where the search spaces are typically too large to be explored in a reasonable amount of time with a reasonable amount of space. Theorem proving, on the other hand, is a more powerful technique than model checking which can deal with an infinite state space. It works by generating a collection of mathematical proof obligations from the system and its specifications and proving these obligations using either proof assistants (such as Coq, Isabelle, PVS, HOL) or using automatic theorem provers (such as SMT solvers). Theorem proving allows us to work on more accurate representations of the system and express any properties, but most proofs have to be done manually which requires time and expertise. Since the binary search tree that we target for verification is a concurrent data structure of unbounded key-value pairs, and processes. The concurrency considerably increases the state space, and we also want to work with a model that's connected to the actual behavior of C programs, while model checking usually works on more abstract models. So, we use the theorem proving approach to formalize the proof of correctness.

In this thesis, we describe the formal verification of the correctness of concurrent binary search tree with fine-grained locking. We implement the BST algorithms in C and use the Verified Software Toolchain(VST) \cite{plfcc} to verify with respect to given specification. The concurrent separation logic (CSL) is employed to reason about behavior of the program implemented in the C programming language. Separation logic (SL) \cite{seplogic} is an extension of Hoare logic \cite{hoare}, way of reasoning about programs using a set of logical rules. SL helps to locally reason about programs that play with pointer data structures, where specifications and proofs of a program component only reflect the chunk of memory that the component uses, not the entire global state of the system. Concurrent separation logic (CSL) is a variant of the separation logic for reasoning about the concurrent programs. The version of CSL formalized in VST is rooted on the work of Hobor et al. \cite{oraclesematic}, but extended with higher-order ghost state in the style of Iris \cite{higherorderghoststate}.
\section{Background}
\label{background}
\subsection{VST and Iris}
We use the Verified Software Toolchain \cite{plfcc}  to prove the correctness of the concurrent binary search tree. VST is a Coq-based proof system that helps to prove separation logic properties of C programs with the assurance that the properties will be preserved in compiled code. First, we write programs in C programming and using CompCert’s \emph{clightgen}, the ASTs for C programs can be generated in Coq. We then state and prove specifications for them using Verifiable C. Verifiable C is a higher-order separation logic for reasoning about the functional correctness of C programs and carries a soundness proof that links high-level specifications to assembly code generated by the CompCert Verified compiler. Verifiable C supports reasoning about two separate kinds of concurrency:  Pthreads-style concurrency with locks (used by implementing blocking data structures that can be fine-grained or coarse-grained ), and concurrency using C11 atomic operations (i.e. lock-free data structures).

@@

$$\inference[\textsf{inv\_alloc}]{}{\triangleright P \vdash \pvs[E] \mathsf{EX}\ i : \mathsf{iname}, \knowInv{i}{P}}$$
$$\inference[\textsf{inv\_open}]{i \in E}{\knowInv{i}{P} \vdash \pvs[E][E \setminus i] \triangleright P * (\triangleright P \wand \pvs[E \setminus i][E] \mathsf{emp})}$$

where invariants are provided in the form of assertion  $\knowInv{i}{P}$, and states that $P$ is maintained as an invariant on the global state with name $i$. The operator%%  $\pvs[][]$
is called ``fancy update" operator,that allows to allocate, open, and close the invariants and the later $\triangleright$ operator is used for impredicativity (i.e. $\knowInv{i}{..\knowInv{i}{P}..}$).
\subsection{Atomic Specifications}
\emph{Linearizability} is one of the most popular traditional correctness condition for concurrent data structures. It is defined in terms of a set of high-level operations, such as insert and remove for a set; if a data structure is linearizable, any operation performed on it tends to happen atomically and in some order that corresponds to the real-time ordering of such operations. For instance, if operation $A$ is completed before operation $B$, then $B$ should logically follow $A$. We can prove the linearizability property of concurrent data structures using \emph{global invariants}. Suppose we have an invariant that said ``there exists a linear history of operations performed on the data structures, stored in some ghost state, and the current  state of the data structure is the one that would result from applying each of those operations in order". Then this directly implies that data structure is linearizable. We can allow a data structure operation to access the content of global invariants, treating that operation as an atomic operation if we are able prove that that operation \emph{appears to execute atomically} at some linearization point. This idea was introduced in the TaDA logic \cite{tada} in the form of \emph{atomic Hoare triples} as shown below, and called as  \emph{Logical atomicity}.    
$$\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$

where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of shared data structure. $P_l$ and $Q_l$ are \emph{local} to each thread 
and do not interfere with global state of a data structure, while $P_p$ and $Q_p$ are public to all threads and must satisfy following rule: $P_p$ must be true at every point from the beginning of the function until the linearization point, for some abstract value $a$, and $Q_p$ must be true at the lineariztion point, but may not still true by the end of the function. So, at the linearization point the operation atomically transitions from $P_p$ to $Q_p$.

VST has encoded an atomic Hoare triple, and provide the way that matches the notation VST uses for normal specifications. Such specification is called \emph{atomic specification}, and can be formalized in Coq as shown in  \ref{atomic_spec}.  This is how we write pre- and postcondition in VST. 
\begin{figure}[htb]
\centering
\begin{verbatim}
Program Definition insert_spec :=
DECLARE _insert
ATOMIC TYPE W OBJ a INVS Ei Eo
WITH ...
PRE [ ... ]
  PROP (...)
  LOCAL (...)
  SEP (P_l) | (P_p)
POST [ ... ]
  PROP ()
  LOCAL ()
  SEP (Q_l) | (Q_p)
\end{verbatim}
\caption{Atomic Specification in VST}
\label{atomic_spec}
\end{figure}
\texttt{W} is the \textsf{TypeTree} representing the type of arguments passed with \texttt{WITH} clause; \texttt{a} is the abstract state for the triple; \texttt{Ei} and \texttt{Eo} are the sets of invariants names inside and outside the triple. The \texttt{PROP} clause describes things that that are true independent of program state, the \texttt{LOCAL} clause describes the values contained in C local variables, and the \texttt{SEP} clause represents the \emph{separating conjunction} (*) of \emph{spatial predicates}, predicates on some part of the memory. In VST, while proving that a function implements an atomic specification, the precondition will contains an $\mathsf{atomic\_shift}$ assertion with the public pre- and postcondition, and the masks inside it. This atomic shift can be accessed through following two rules:
$$\inference[\textsf{atomic\_commit}]{\forall a, R * P_p\ a \Rrightarrow \mathsf{EX}\ y,\ Q_p\ a\ y * R'\ y}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
$$\inference[\textsf{atomic\_rollback}]{\forall a, R * P_p\ a \Rrightarrow P_p\ a * R'}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R'}$$
During commit, we must provide resources $R$ which, in combination with the public precondition $P_p$, allow us to prove the public postcondition $Q_p$. Then, we gain access to an assertion $Q$ required by the postcondition of the function, and leftover resources $R'$. During rollback, we provide resources $R$ which, in combination with the public precondition, allow us to reestablish the public precondition; we then regain the atomic shift back in the proof, as well as leftover resources $R'$. The rollback is specially used to learn some relationship between pieces of information (e.g. ghost state) stored in the public precondition, while the commit is used to prove the public postcondition from the precondition, and obtain an assertion $Q$. To complete the proof of any function's specification, we must always perform commit to obtain $Q$; we can perform any number of roolbacks before that point, but after that point we lose the atomic shift and no access to the public precondition.

\section{Safety Proofs}
\label{safety}
\begin{figure}[htp]
\begin{subfigure}[t]{\textwidth}
\begin{lstlisting}[language = C]
typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
typedef struct tree_t **treebox;

@@

\subsection{Specification and Verification}
We begin the verification by defining a predicate $\mathsf{nodebox\_rep}(p)$ that describes the concrete representation of the tree, where $p$ is the pointer to the root node. The $\texttt{nodebox\_rep}$ predicate ties together $\texttt{lock\_inv}$ assertion for each recursive lock as follows: 
\begin{align*}
 &\mathsf{nodebox\_rep}_{\pi}(p) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \exists\ pa\ pb\ \ tp\mapsto\ (pa * pb)\ *\  \mathsf{nodebox\_rep}_{0.5}(pa)\ *\ \mathsf{nodebox\_rep}_{0.5}(pb)  \end{align*}

@@

The resource invariant $R$ need to be \emph{exclusive}, i.e, it can only hold once in any given state. This allows us to know that if the current thread holds the invariants, it also hold the lock.

\section{Correctness Proofs}
\label{correctness}
In the previous section, we showed that how to prove concurrent data-structure is thread safe, but not functionally correct. To prove the correctness, our threads need to be able to record the information about the actions they have performed on the shared state, instead of sealing all knowledge of the shared data structure inside the lock invariant. We can accomplish this with ghost variables, a simple form of auxiliary state. Aside from the \emph{permission} on shared resources held by each thread that we used in the safety proofs, the verification of correctness properties also involves the \emph{ghost state}. In VST, any Coq type can be used as ghost state, as long as we can describe what happens when two elements of that type are joined together. We can use own predicate in VST to represent ghost state assertion as follows: $\mathsf{own}\ g\ a\ \mathit{pp}$, where $g$ is a \emph{ghost name}, $a$ is the value associated with $g$ which can be of any type, and $\mathit{pp}$ is a separation logic predicate.

@@

\subsection{Specification of the algorithm}
In addition to program state, the verification of concurrent program involves two aspects which do not appear in actual C program: the \emph{shares} on shared resources and \emph{ghost state} described in Section \ref{safety}. We need to extend our specification in Section\ref{safety} with \emph{ghost state} to prove the functional correctness.  

@@

The range for root node is always $(-\infty,+\infty)$ and then, we can calculate the range for all nodes in the tree by enforcing each node to propagates appropriate bounds. For the left child, range will be lower-bound for the node (lower-bound) and the key in the node (upper-bound), while for the right child range will be a key in the node (lower-bound) and upper-bound for the node (upper-bound). This properties in each node help us to prove that the thread have changed the state of data-structure to the valid state. For instance, we can confirm that a key has been inserted in the right place if a key is inside the range of the node. Suppose we want to insert a key $38$ in \ref{range_bst}, then we will locate \emph{right} child of $35$ as the right place for $38$. During insertion, we can confirm that the node is indeed right place for $38$ by checking that $38$ is in the range (35,40). We introduce the $range$ in our proof using the ghost variables. This $ranges$ along with other information about the node represent the abstract state of the data-structure. We have used two kind of ghost state in our proof: \emph{per-node} ghost state and \emph{over-all} ghost state.\\

@@

\textbf{Over-all Ghost State}: We have constructed the per-node ghost state for each node, using unique ghost variable for each node. But if we are given a ghost variable (\emph{gname}) to access the resources in the node labeled with that variable, we do not any clue about whether or not that variable exists in the tree. So, we need another kind of ghost state, which represents the a set of all per-node ghost variable. Again We can use ``reference" pattern to define this ghost state:
\begin{verbatim}
Definition ghost_ref g r1 := ghost_reference(P := set_PCM) r1 g.
Definition in_tree g g1 := EX sh: share, ghost_part(P :=set_PCM) sh (Ensembles.Singleton g1) g.
\end{verbatim}
We use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part of the atomic specification respectively. The global invariant, which we introduced in our atomic specification as $\mathsf{tree\_rep}$ predicate above, ties together $\texttt{public\_half}$ and $\texttt{ghost\_ref}$ as follows:
\begin{align*}\mathsf{tree\_rep}(T,g) \triangleq [\ \forall n. \mathsf{public\_half\textsubscript{gn}} \ (\mathit{range}_n* \mathit{ghost\_info}_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
 Later while writing atomic specification, we will use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part respectively. Once we open \emph{global invariant}, we get access to $\mathsf{ghost\_ref}$ which will be used with $\mathsf{in\_tree}$ to establish the fact that node actually exists in the tree. We proved the following lemma for this:
$$
Lemma\ \mathbf{\ node\_exist\_in\_tree : }\ \forall\ g\ s\ g\_in,\  \mathbf{in\_tree}\ g\ g\_in  * \mathbf{ghost\_ref}\ g\ s\ \Rightarrow\ g\_in\ \in s .
$$
\subsubsection{Combining all pieces in atomic specification}
We combine pieces of information discussed above, \emph{recursive lock invariant} from \ref{safety} and \emph{ghost states}, to write the atomic specification. The private pre- and postcondition of the atomic specification is given by defining the predicate $\mathsf{nodebox\_rep}(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and over-all ghost state is represented by ghost name $g$. The $\mathsf{nodebox\_rep}$ predicate ties together $\mathsf{my\_half}$ part of per-node ghost state and $\mathsf{in\_tree}$ of over-all ghost state inside the $\mathsf{lock\_inv}$ assertion of each node's lock as follows:
\begin{align*}
    &\mathsf{nodebox\_rep}_{\pi}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \ghost{\mathsf{my\_half}(range,info)}{g\_root}\ *\ \ghost{\mathsf{in\_tree}\ (g\_root)}{g}\ *\ \exists\ pa\ pb\ ga\ gb,\ tp\mapsto\ (pa * pb)\ *\  \mathsf{nodebox\_rep}_{0.5}(pa,g,ga) \\& *\ \mathsf{nodebox\_rep}_{0.5}(pb,g,gb))\end{align*}

The $\mathsf{nodebox\_rep}$ is a predicate describing the local information about the node almost similar to the predicate we defined for the safety proof in Section \ref{safety}. In this case, the lock invariant of recursive lock have, along with the information about the children node,$\mathsf{my\_half}$ predicate which holds the partial information describing the current thread's contribution to the shared state.

>>>>

\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{iris}
%\usepackage{lstlangcoq}
\renewcommand{\lstlistingname}{Figure}

@@

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\treerep}{\ensuremath{\mathsf{bst}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{bst\_ref}}}

\newcommand{\ignore}[1]{}

@@

\title{Verifying Fine-Grained and Lock-Free Binary Search Trees}
\author{Roshan Sharma}
\author{Shengyi Wang}
\author{Lennart Beringer}
\author{William Mansky}

@@

\catcode`\@=11
\section{Introduction}
\input{intro.tex}

\section{Background}
\label{background}
\subsection{VST and Iris}
We use the Verified Software Toolchain \cite{plfcc}  to prove the correctness of the concurrent binary search tree. VST is a Coq-based proof system that helps to prove separation logic properties of C programs with the assurance that the properties will be preserved in compiled code. First, we write programs in C programming and using CompCert’s \emph{clightgen}, the ASTs for C programs can be generated in Coq. We then state and prove specifications for them using Verifiable C. Verifiable C is a higher-order separation logic for reasoning about the functional correctness of C programs and carries a soundness proof that links high-level specifications to assembly code generated by the CompCert Verified compiler. Verifiable C supports reasoning about two separate kinds of concurrency:  Pthreads-style concurrency with locks (used by implementing blocking data structures that can be fine-grained or coarse-grained ), and concurrency using C11 atomic operations (i.e. lock-free data structures).

@@

\begin{mathpar}
\inference[\textsf{inv\_alloc}]{}{\triangleright P \vdash \pvs[E] \mathsf{EX}\ i : \mathsf{iname}, \knowInv{i}{P}}

\inference[\textsf{inv\_open}]{i \in E}{\knowInv{i}{P} \vdash \pvs[E][E \setminus i] \triangleright P * (\triangleright P \wand \pvs[E \setminus i][E] \mathsf{emp})}
\end{mathpar}

where invariants are provided in the form of assertion $\knowInv{i}{P}$, and states that $P$ is maintained as an invariant on the global state with name $i$. The operator%%  $\pvs[][]$
is called ``fancy update" operator,that allows to allocate, open, and close the invariants and the later $\triangleright$ operator is used for impredicativity (i.e. $\knowInv{i}{..\knowInv{i}{P}..}$).

\subsection{Atomic Specifications}
\label{atomic}
For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. For instance, the specification $$\forall s.\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution. %maybe something about linking ghost state, which parts are real vs. ghost, accessing invariants around atomic triple

VST has encoded an atomic Hoare triple, and provide the way that matches the notation VST uses for normal specifications. Such specification is called \emph{atomic specification}, and can be formalized in Coq as shown in  \ref{atomic-spec}.  This is how we write pre- and postcondition in VST. 
\begin{figure}[htb]
\centering
\begin{verbatim}
Program Definition insert_spec :=
DECLARE _insert
ATOMIC TYPE W OBJ a INVS Ei Eo
WITH ...
PRE [ ... ]
  PROP (...)
  LOCAL (...)
  SEP (P_l) | (P_p)
POST [ ... ]
  PROP ()
  LOCAL ()
  SEP (Q_l) | (Q_p)
\end{verbatim}
\caption{Atomic Specification in VST}
\label{atomic-spec}
\end{figure}
\texttt{W} is the \textsf{TypeTree} representing the type of arguments passed with \texttt{WITH} clause; \texttt{a} is the abstract state for the triple; \texttt{Ei} and \texttt{Eo} are the sets of invariants names inside and outside the triple. The \texttt{PROP} clause describes things that that are true independent of program state, the \texttt{LOCAL} clause describes the values contained in C local variables, and the \texttt{SEP} clause represents the \emph{separating conjunction} (*) of \emph{spatial predicates}, predicates on some part of the memory. In VST, while proving that a function implements an atomic specification, the precondition will contains an $\mathsf{atomic\_shift}$ assertion with the public pre- and postcondition, and the masks inside it. This atomic shift can be accessed through following two rules:
$$\inference[\textsf{atomic\_commit}]{\forall a, R * P_p\ a \Rrightarrow \mathsf{EX}\ y,\ Q_p\ a\ y * R'\ y}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
$$\inference[\textsf{atomic\_rollback}]{\forall a, R * P_p\ a \Rrightarrow P_p\ a * R'}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R'}$$
During commit, we must provide resources $R$ which, in combination with the public precondition $P_p$, allow us to prove the public postcondition $Q_p$. Then, we gain access to an assertion $Q$ required by the postcondition of the function, and leftover resources $R'$. During rollback, we provide resources $R$ which, in combination with the public precondition, allow us to reestablish the public precondition; we then regain the atomic shift back in the proof, as well as leftover resources $R'$. The rollback is specially used to learn some relationship between pieces of information (e.g. ghost state) stored in the public precondition, while the commit is used to prove the public postcondition from the precondition, and obtain an assertion $Q$. To complete the proof of any function's specification, we must always perform commit to obtain $Q$; we can perform any number of roolbacks before that point, but after that point we lose the atomic shift and no access to the public precondition.

\ignore{\section{Safety Proofs} %maybe cut this, or roll it into the next section
\label{safety}
\begin{figure}[htp]
\begin{subfigure}[t]{\textwidth}
\begin{lstlisting}[language = C]
typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
typedef struct tree_t **treebox;

@@

\subsection{Specification and Verification}
We begin the verification by defining a predicate $\nodeboxrep(p)$ that describes the concrete representation of the tree, where $p$ is the pointer to the root node. The $\texttt{nodebox\_rep}$ predicate ties together $\texttt{lock\_inv}$ assertion for each recursive lock as follows: 
\begin{align*}
 &\nodeboxrep_{\pi}(p) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \exists\ pa\ pb\ \ tp\mapsto\ (pa * pb)\ *\  \nodeboxrep_{0.5}(pa)\ *\ \nodeboxrep_{0.5}(pb)  \end{align*}

@@

The resource invariant $R$ need to be \emph{exclusive}, i.e, it can only hold once in any given state. This allows us to know that if the current thread holds the invariants, it also hold the lock. 
}

\section{Correctness Proofs}
\label{correctness}
As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
\begin{mathpar}
\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle

\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle

\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{delete}\ t\ k)\rangle
\end{mathpar}
In our case, insert, lookup, and delete are simple Gallina functions on a functional tree data structure. The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract tree $t$.

In the following sections, we describe the construction of these predicates, and the proofs of the operations, for BST implementations based on hand-over-hand locking and lock-free atomic operations.

%look at some of the Iris papers in CPP, etc. for how we should describe the proofs

%%%%%%%%%%%%%%%%%%%%%

\ignore{In the previous section, we showed that how to prove concurrent data-structure is thread safe, but not functionally correct. To prove the correctness, our threads need to be able to record the information about the actions they have performed on the shared state, instead of sealing all knowledge of the shared data structure inside the lock invariant. We can accomplish this with ghost variables, a simple form of auxiliary state. Aside from the \emph{permission} on shared resources held by each thread that we used in the safety proofs, the verification of correctness properties also involves the \emph{ghost state}. In VST, any Coq type can be used as ghost state, as long as we can describe what happens when two elements of that type are joined together. We can use own predicate in VST to represent ghost state assertion as follows: $\mathsf{own}\ g\ a\ \mathit{pp}$, where $g$ is a \emph{ghost name}, $a$ is the value associated with $g$ which can be of any type, and $\mathit{pp}$ is a separation logic predicate.}

@@

\subsection{Specification of the hand-over-hand BST}
%In addition to program state, the verification of concurrent program involves two aspects which do not appear in actual C program: the \emph{shares} on shared resources and \emph{ghost state} described in Section \ref{safety}. We need to extend our specification in Section\ref{safety} with \emph{ghost state} to prove the functional correctness.

To verify our tree implementation, we need to instantiate $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure), and show how they interact to access and update the state of the data structure for each call. In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure. Our overall approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$. We also need to make sure that our lock invariants support the hand-over-hand pattern.

@@

The range for root node is always $(-\infty,+\infty)$ and then, we can calculate the range for all nodes in the tree by enforcing each node to propagates appropriate bounds. For the left child, range will be lower-bound for the node (lower-bound) and the key in the node (upper-bound), while for the right child range will be a key in the node (lower-bound) and upper-bound for the node (upper-bound). This properties in each node help us to prove that the thread have changed the state of data-structure to the valid state. For instance, we can confirm that a key has been inserted in the right place if a key is inside the range of the node. Suppose we want to insert a key $38$ in \ref{range_bst}, then we will locate \emph{right} child of $35$ as the right place for $38$. During insertion, we can confirm that the node is indeed right place for $38$ by checking that $38$ is in the range (35,40). We introduce the $range$ in our proof using the ghost variables. This $ranges$ along with other information about the node represent the abstract state of the data structure. We have used two kind of ghost state in our proof: \emph{per-node} ghost state and \emph{overall} ghost state.\\

@@

\textbf{Overall Ghost State}: We have constructed the per-node ghost state for each node, using unique ghost variable for each node. But if we are given a ghost variable (\emph{gname}) to access the resources in the node labeled with that variable, we do not any clue about whether or not that variable exists in the tree. So, we need another kind of ghost state, which represents the a set of all per-node ghost variable. Again We can use ``reference" pattern to define this ghost state:
\begin{verbatim}
Definition ghost_ref g r1 := ghost_reference(P := set_PCM) r1 g.
Definition in_tree g g1 := EX sh: share, ghost_part(P :=set_PCM) sh (Ensembles.Singleton g1) g.
\end{verbatim}
We use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part of the atomic specification respectively. The global invariant, which we introduced in our atomic specification as $\treerep$ predicate above, ties together $\texttt{public\_half}$ and $\texttt{ghost\_ref}$ as follows:
\begin{align*}\treerep(T,g) \triangleq [\ \forall n. \mathsf{public\_half\textsubscript{gn}} \ (\mathit{range}_n* \mathit{ghost\_info}_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
 Later while writing atomic specification, we will use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part respectively. Once we open \emph{global invariant}, we get access to $\mathsf{ghost\_ref}$ which will be used with $\mathsf{in\_tree}$ to establish the fact that node actually exists in the tree. We proved the following lemma for this:
$$
Lemma\ \mathbf{\ node\_exist\_in\_tree : }\ \forall\ g\ s\ g\_in,\  \mathbf{in\_tree}\ g\ g\_in  * \mathbf{ghost\_ref}\ g\ s\ \Rightarrow\ g\_in\ \in s .
$$
\subsubsection{Combining all pieces in atomic specification}
We combine pieces of information discussed above, \emph{recursive lock invariant} from \ref{safety} and \emph{ghost states}, to write the atomic specification. The private pre- and postcondition of the atomic specification is given by defining the predicate $\nodeboxrep(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and over-all ghost state is represented by ghost name $g$. The $\nodeboxrep$ predicate ties together $\mathsf{my\_half}$ part of per-node ghost state and $\mathsf{in\_tree}$ of over-all ghost state inside the $\mathsf{lock\_inv}$ assertion of each node's lock as follows:
\begin{align*}
    &\nodeboxrep_{\pi}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \ghost{\mathsf{my\_half}(range,info)}{g\_root}\ *\ \ghost{\mathsf{in\_tree}\ (g\_root)}{g}\ *\ \exists\ pa\ pb\ ga\ gb,\ tp\mapsto\ (pa * pb)\ *\  \nodeboxrep_{0.5}(pa,g,ga) \\& *\ \nodeboxrep_{0.5}(pb,g,gb))\end{align*}

The $\nodeboxrep$ is a predicate describing the local information about the node almost similar to the predicate we defined for the safety proof in Section \ref{safety}. In this case, the lock invariant of recursive lock have, along with the information about the children node,$\mathsf{my\_half}$ predicate which holds the partial information describing the current thread's contribution to the shared state.

The public pre- and postcondition of the atomic specification is given by defining the predicate $\treerep(T,g)$ that describes the abstract state of the binary search tree $T$  in terms of collection of ghost states each describing the information about separate node in the tree.. The $\treerep$ predicate ties together $\mathsf{public\_half}$ part of per-node ghost states and $\mathsf{ghost\_ref}$ of over-all ghost state as follows: \begin{align*}&\mathsf{ghost\_tree\_rep}(tg,range, g, g\_current) \triangleq\ \exists\ k, v, ga, gb\ \in tg,\ghost{\mathsf{public\_half}\ (range, Some(k,v,ga,gb))}{g\_current}\\& *\ \mathsf{ghost\_tree\_rep}(left(tg),(left(range),k),g,ga)\ *\ \mathsf{ghost\_tree\_rep}(right(tg),(k, right(range)),g,gb) \end{align*}
\begin{align*}\treerep(T,g,g\_root) \triangleq \ \exists\ (tg:ghost\_tree),\ &\mathsf{pure\_tree}(tg)\ =\ T\ \land\ \mathsf{ghost\_tree\_rep}(tg,(-\infty, +\infty),g,g\_root)&\\ *\ \ghost{\mathsf{ghost\_ref}\ (\mathsf{find\_ghost\_set}(tg))}{g}\end{align*}
Where $tg$ is called \emph{ghost tree}, which can be existentially quantified, extend the node in pure tree $T$ with actual pointer and ghost names for child nodes.

The atomic specification for any binary search tree's operation \texttt{op} can be written as follows:
\begin{align*}\forall t.\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t,g)\rangle\ \texttt{op}(p)\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t', g)\rangle
\end{align*}
